\vskip-0.6in Trade Agreements in the Shadow of Lobbying\protect\footnote{Formerly circulated under the title ``Trade Agreements, Lobbying and Separation of Powers''}

### 2016-02-02 draft.tex
final proofread before submission

\begin{quote}
{This paper presents a model of international trade agreements in which the executive branches of each government negotiate agreements while the legislative branches, subject to political pressure from firms, can disrupt them. Lobbying is in the style of Grossman and Helpman (1994) with a new feature: all actors face uncertainty arising from the complexity of the legislative process. I demonstrate that the lower the executives set tariffs in a trade agreement, the more effort lobbies put forth to prevent its ratification. Thus trade agreements act as a domestic political commitment device: executives set relatively high tariffs to discourage lobbying and increase the chance that the agreement will be ratified. \textbf{This finding sheds light on the empirical puzzle surrounding governments' welfare weights in the Grossman and Helpman (1994) model, while the addition of political uncertainty adds important realism so that the model presented here can speak to important trade policy-making events such as the four-year delay in the ratification of the U.S.-Korea Free Trade Agreement.}

@@

\textbf{Assuming for now that lobbies represent only import-competing interests, lobbies in this environment would like to prevent or derail a trade agreement that would reduce their tariff protection. They work toward this end by exerting effort to convince their legislatures to vote against the trade agreement but face some uncertainty about whether their efforts will result in a voting outcome consistent with their interests. I show that the probability that the trade agreement is derailed increases in lobbying effort and decreases in the trade agreement tariff. Further, the lobby decreases its effort as the tariff it would face under the trade agreement rises: it has less to lose the higher is the trade agreement tariff. When forming the trade agreement tariff, the executives trade off the welfare they receive under the agreement with the probability that the agreement actually remains in force. Ultimately, I find that observed lobbying effort may often be low because the executive branches of the governments set trade agreement tariffs sufficiently high to reduce political pressure on the legislatures. I also show that other examples of realism are gained when political uncertainty is added to the model: for instance, the model rationalizes the four-year delay and eventual ratification of the U.S.-Korea Free Trade Agreement.}

@@

To the best of my knowledge, this is the first attempt to model lobbying specifically aimed at derailing trade agreements. Although this kind of politically-driven failure is not rare, prior models have not allowed exploration of the endogenous dynamics behind them. Careful consideration of the political process seems to be important for understanding why governments so often fail to cooperate. The recent ratification of the U.S.-Korea Free Trade Agreement is a case in point. The agreement, as originally negotiated in 2007, was not able to be ratified in the United States until President Obama gained concessions for U.S. automakers, after which they removed their objections and the agreement was ratified by Congress in 2011 and entered into force in 2012. In a certain world, this ratification failure should not have occurred, and it was remedied by making the trade agreement more appealing to a key lobby and thus reducing its lobbying effort. \footnote{See \url{http://www.ustr.gov/trade-agreements/free-trade-agreements/korus-fta}. In this case, automakers are both importers and exporters, and the concessions were concentrated on market access for auto exports in South Korea. For simplicity, the model presented herein focuses on lobbying by an import-competing industry in a model of interindustry trade; the central insights do not depend on this structure and can be extended to the case of multiple lobbies and/or intraindustry trade. In particular, adding an export lobby in the simplest case of no uncertainty predicts that import lobbies will remain disengaged while the executive will be able to reduce tariffs by recruiting support from exporters. The Panama Trade Promotion Act seems to follow this pattern, where there was virtually no lobbying against, significant lobbying in favor, and numerous industries receiving import tariff protection.\label{fn:lobby}}

@@

\textbf{The addition of a rich structure of government}, tailored carefully to the trade policy making process, helps shed light on the empirical puzzle surrounding the PFS model. Empirical tests of the model\footnote{cfr. Goldberg and Maggi (1999); Gawande and Bandyopadhyay (2000); Mitra, Thomakos and Ulubasoglu (2002); McCalman (2004).\label{fn:gh_lobby}} have consistently found the weight governments place on social welfare to be many times that which they place on lobbying effort, while estimates indicate that the deadweight losses caused by trade distortions are several orders of magnitude larger than lobbying expenditures.\footnote{Feenstra (1992) assembles estimates from the mid-1980s that place a floor of $\$$8 billion per year on the deadweight losses from protection, whereas Bombardini and Trebbi (2012) find that total annual lobbying expenditure on trade issues in 1999-2001 when this data first became available was about $\$$200 million.} This raises the question: how could governments value social welfare so highly, yet grant these quantities of protection at such a low price?

@@

In the non-unitary government model, these estimates and the earlier-cited stylized facts become consistent with each other. The ``case of the missing contributions,'' as Gawande and Hoekman (2006) refer to it, turns out to be well explained by an equilibrium phenomenon arising from the agenda-setting power of the executive branch. In particular, \textbf{in this model, there is inconsistency for the approximately 15$\%$ of sectors that receive protection but apparently put forth no lobbying effort. In this framework, even a welfare-maximizing executive and lobbying effort at zero can be reconciled with high tariff levels.}

@@

\textbf{After this}, the uncertainty about the the median legislator's identity is resolved. This is modeled as the realization of a random variable $\ta$, the value of which all players observe simultaneously. The ``break stage'' concludes with the legislature making a choice to abide by the agreement or to break the agreement.

@@

The home executive's welfare is specified as follows:
\[
  W_E = \mathit{CS}_X(\tau) + \ga_E \cdot \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau)
\]
Note that this is identical to the welfare function for the legislature aside from the weight $\ga_E$ the executive places on the profits of the import industry, which is not a function of lobbying effort. \textbf{Section~\ref{sec:choices} discusses these modeling choices in depth.}

@@

\textbf{Finally, I assume that an external authority can ensure enforcement of the agreement in the one-shot game analyzed here.}\footnote{Buzard (\textbf{2015}) extends the analysis to a repeated-game context and can therefore remove this restriction.}

@@

Notice that, aside from the endogeneity of the weight the legislature places on the lobbying industry's profits, \textbf{Equation~\ref{eq:ml}} is precisely the \textit{deus ex machina} government objective function popularized by Baldwin (1987) that is commonly employed in the literature on the political economy of trade agreements. \textbf{Instead of the weight on profits being constant as in the exogenous case or necessarily linear in contributions as in the PFS case, this form allows for both endogeneity and non-linearity.}

Since trade policies are overwhelmingly determined within the context of trade agreements, it is useful to have a framework to bring together the endogenous political pressure of PFS-style modeling with the trade agreements approach. The formulation in Equation~\ref{eq:ml} is intended as a bridge between the two. See Buzard (\textbf{2016}) for an example of how this form can be used to integrate the analysis of endogenous political activity and exogenous shocks to the political welfare of governments.

\textbf{In addition, letting the weight the median legislator places on the profits of the import-competing industry $\ga(e,\ve)$ be affected by the level of lobbying effort as well a random variable $\ta$} allows for the analysis of endogenous lobbying to be integrated with the standard political economy shocks of the trade agreements literature. Modeling the objective function so closely on the standard in the trade agreements literature allows for direct comparisons to the large extant body of work that studies exogenous shocks only, revealing cleanly the effects of the addition of endogenous lobbying---and if one wants---this different kind of political uncertainty.

@@

Therefore, in line with both the theoretical and empirical literature, I make the following, \textbf{final} assumption \textbf{on $\ga(e,\ta)$}:

@@

\textbf{The results apply beyond separation-of-powers presidential systems because the presence of veto points is what is essential for the dynamic posited here to occur. Although the number of formal veto points tends to be lower in, for instance, parliamentary systems, veto points are typically not absent altogether (Henisz and Mansfield 2006). In reality, constraints on the executive are almost always present in democratic regimes, although they may operate in different ways, e.g. over questions of government dissolution and formation in parliamentary regimes. It is important to consider that constraints may be formal or informal, and they may come from legislative, regulatory, judicial, sub-national or even other executive bodies.
}

@@

\section{Main Results}
\label{sec:main}
To understand how the executives optimally structure trade agreements, we must first examine the incentives of the lobbies and how the legislatures make decisions regarding breach of the trade agreement, including how trade-war tariffs are set. The symmetric structure of the model permits restriction of attention to the home country.

@@

In the event of a trade war, the lobby chooses its effort $e_n$ given this tariff-setting behavior by maximizing its profits net of effort: $\pi\left(\tau^n\left(\ga\left(e_n,\ve_n\right)\right)\right) - e_n$. (Note this is Equation~\ref{eq:lv} simplified by the resolution of uncertainty over the legislature's decision on the trade agreement). This implies a first order condition \textbf{for the lobby} of
\begin{equation}
	\bm{\frac{\partial \pi(\tau^n)}{\partial \tau^n}\frac{\partial \tau^n}{\partial \ga} \frac{\partial \ga}{\partial e_n} = 1}
  \label{eq:lobtw}
\end{equation}
That is, at this stage, the lobby chooses the level of effort that equates its expected marginal increase in profits with its marginal payment. Further details about the political weighting function $\ga$ and assumed structure of uncertainty are provided in the next section.

@@

\bigskip
\subsection{To Break or Not to Break?}
\label{sec:break}
With trade-war decisions specified, we can proceed to analyze the incentives regarding the legislature's decision to uphold or break the trade agreement. I assume that the form of the political weighting function is the same \textbf{regardless of the break decision}, but the model could be extended to allow the two to differ. What is in principle different are the lobbying efforts toward influencing the break decision, denoted $e_b$ at this stage to distinguish them from those that are made to influence the Nash tariff, $e_n$. The realization of uncertainty about the median legislator's identity will similarly be denoted $\ve_b$ so as to distinguish it from $\ve_n$ at the trade-war stage.

@@

The executives' incentive to raise tariffs in order to reduce lobbying effort helps to explain the empirical finding in the Protection for Sale literature that levels of protection and associated deadweight losses are too high relative to lobbying expenditure given the high estimates for governments' weighting of social welfare. I have shown that this serves to mediate the relationship between preferences for contributions relative to social welfare and the tariffs that are provided. The intuition is clear: the observed lobbying expenditure levels may in fact be low \textit{because} tariffs have been raised sufficiently high to prevent political pressure and the increased risk of a costly trade disruption it engenders.

>>>>

\begin{quote}
{This paper presents a model of international trade agreements in which the executive branches of each government negotiate agreements while the legislative branches, subject to political pressure from firms, can disrupt them. Lobbying is in the style of Grossman and Helpman (1994) with a new feature: all actors face uncertainty arising from the complexity of the legislative process. I demonstrate that the lower the executives set tariffs in a trade agreement, the more effort lobbies put forth to prevent its ratification. Thus trade agreements act as a domestic political commitment device: executives set relatively high tariffs to discourage lobbying and increase the chance that the agreement will be ratified. \textbf{The model sheds light on the empirical puzzle surrounding governments' welfare weights in the Grossman and Helpman (1994) model and can speak to important trade policy-making events such as the four-year delay in the ratification of the U.S.-Korea Free Trade Agreement.}

@@

\textbf{Assuming that lobbies represent only import-competing interests, lobbies in this environment would like to prevent or derail a trade agreement that would reduce their tariff protection. They work toward this end by exerting effort to convince their legislatures to vote against the trade agreement but face some uncertainty about whether their efforts will result in a voting outcome consistent with their interests. I show that the probability that the trade agreement is derailed increases in lobbying effort and decreases in the trade agreement tariff. Further, the lobby decreases its effort as the tariff it would face under the trade agreement rises: it has less to lose the higher is the trade agreement tariff. When forming the trade agreement tariff, the executives therefore trade off the welfare they receive under the agreement with the probability that the agreement actually remains in force. Ultimately, I find that observed lobbying effort may often be low because the executive branches of the governments set trade agreement tariffs sufficiently high to reduce political pressure on the legislatures. There are other ways in which realism is gained when political uncertainty is added to the model: for instance, the model rationalizes the four-year delay and eventual ratification of the U.S.-Korea Free Trade Agreement.}

@@

To the best of my knowledge, this is the first attempt to model lobbying specifically aimed at derailing trade agreements. Although this kind of politically-driven failure is not rare, prior models have not allowed exploration of the endogenous dynamics behind them. Careful consideration of the political process seems to be important for understanding why governments so often fail to cooperate. The recent ratification of the U.S.-Korea Free Trade Agreement is a case in point. The agreement, as originally negotiated in 2007, was not able to be ratified in the United States until President Obama gained concessions for U.S. automakers, after which they removed their objections and the agreement was ratified by Congress in 2011 and entered into force in 2012. In a certain world, this ratification failure should not have occurred, and it was remedied by making the trade agreement more appealing to a key lobby and thus reducing its lobbying effort.\footnote{See \url{http://www.ustr.gov/trade-agreements/free-trade-agreements/korus-fta}. In this case, automakers are both importers and exporters, and the concessions were concentrated on market access for auto exports in South Korea. For simplicity, the model presented herein focuses on lobbying by an import-competing industry in a model of interindustry trade; the central insights do not depend on this structure and can be extended to the case of multiple lobbies and/or intraindustry trade. In particular, adding an export lobby in the simplest case of no uncertainty predicts that import lobbies will remain disengaged while the executive will be able to reduce tariffs by recruiting support from exporters. The Panama Trade Promotion Act seems to follow this pattern, where there was virtually no lobbying against, significant lobbying in favor, and numerous industries receiving import tariff protection.\label{fn:lobby}}

@@

\textbf{The addition of a rich structure of government}, tailored carefully to the trade policy making process, also helps shed light on the empirical puzzle surrounding the PFS model. Empirical tests of the model\footnote{cfr. Goldberg and Maggi (1999); Gawande and Bandyopadhyay (2000); Mitra, Thomakos and Ulubasoglu (2002); McCalman (2004).\label{fn:gh_lobby}} have consistently found the weight governments place on social welfare to be many times that which they place on lobbying effort, while estimates indicate that the deadweight losses caused by trade distortions are several orders of magnitude larger than lobbying expenditures.\footnote{Feenstra (1992) assembles estimates from the mid-1980s that place a floor of $\$$8 billion per year on the deadweight losses from protection, whereas Bombardini and Trebbi (2012) find that total annual lobbying expenditure on trade issues in 1999-2001 when this data first became available was about $\$$200 million.} This raises the question: how could governments value social welfare so highly, yet grant these quantities of protection at such a low price?

@@

In the non-unitary government model, these estimates and the earlier-cited stylized facts become consistent with each other. The ``case of the missing contributions,'' as Gawande and Hoekman (2006) refer to it, turns out to be well explained by an equilibrium phenomenon arising from the agenda-setting power of the executive branch. In particular, \textbf{in this model, there is no inconsistency for the approximately 15$\%$ of sectors that receive protection but apparently put forth no lobbying effort. In this framework, even a welfare-maximizing executive and lobbying effort at zero can be reconciled with high tariff levels.}

@@

\textbf{After this}, the uncertainty about the the median legislator's identity is resolved. This is modeled as the realization of a random variable $\ta_b$, the value of which all players observe simultaneously. The ``break stage'' concludes with the legislature making a choice to abide by the agreement or to break the agreement.

@@

The home executive's welfare is specified as follows:
\[
  W_E = \mathit{CS}_X(\tau) + \ga_E \cdot \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau)
\]
Note that this is identical to the welfare function for the legislature aside from the weight $\ga_E$ the executive places on the profits of the import industry, which is not a function of lobbying effort. \textbf{Section~\ref{sec:choices} contains an in-depth discussion of these modeling choices.}

@@

\textbf{Finally, I assume that an external authority can ensure enforcement of the agreement in the one-shot game analyzed here.}\footnote{Buzard (\textbf{2015}) considers a similar model in a repeated-game context and can therefore remove this restriction.}

@@

Now, notice that, aside from the endogeneity of the weight the legislature places on the lobbying industry's profits, \textbf{Equation~\ref{eq:ml}} is precisely the \textit{deus ex machina} government objective function popularized by Baldwin (1987) that is commonly employed in the literature on the political economy of trade agreements. \textbf{Instead of the weight on profits being constant as in the exogenous case or necessarily linear in contributions as in the PFS case, this form allows for both endogeneity and non-linearity.}

Since trade policies are overwhelmingly determined within the context of trade agreements, it is useful to have a framework to bring together the endogenous political pressure of PFS-style modeling with the trade agreements approach. The formulation in Equation~\ref{eq:ml} is intended as a bridge between the two. \textbf{See Buzard (\textbf{2016}) for an example of how letting the weight the median legislator places on the profits of the import-competing industry $\ga(e,\ve)$ be affected by the level of lobbying effort as well a random variable $\ta$ can integrate both endogenous political activity and exogenous shocks into the analysis of questions regarding optimal trade agreement design and the political welfare of governments.}

Modeling the objective function so closely on the standard in the trade agreements literature allows for direct comparisons to the large extant body of work that studies exogenous shocks only, revealing cleanly the effects of the addition of endogenous lobbying---and if one wants---this different kind of political uncertainty.

@@

Therefore, in line with both the theoretical and empirical literature, I make \textbf{one more} assumption \textbf{on $\ga(e,\ta)$}:

@@

\textbf{We can see that the results apply beyond separation-of-powers presidential systems because the presence of veto points is what is essential for the dynamic posited here to occur. Although the number of formal veto points tends to be lower in, for instance, parliamentary systems, veto points are typically not absent altogether, at least in democracies (Henisz and Mansfield 2006). In reality, constraints on the executive are almost always present in democratic regimes, although they may operate in different ways, e.g. over questions of government dissolution and formation in parliamentary regimes. It is important to consider that constraints may be formal or informal, and they may come from legislative, regulatory, judicial, sub-national or even other executive bodies.
}

@@

\section{Main Results}
\label{sec:main}
To understand how the executives optimally structure trade agreements, we must first examine the incentives of the lobbies and how the legislatures make decisions regarding breach of the trade agreement, including how trade-war tariffs are set. The symmetric, separable structure of the model permits restriction of attention to the home country.

@@

In the event of a trade war, the lobby chooses its effort $e_n$ given this tariff-setting behavior by maximizing its profits net of effort: $\pi\left(\tau^n\left(\ga\left(e_n,\ve_n\right)\right)\right) - e_n$. (Note this is Equation~\ref{eq:lv} simplified by the resolution of uncertainty over the legislature's decision on the trade agreement). This implies a first order condition \textbf{for the lobby} of
\begin{equation}
	\bm{\frac{\partial \pi(\tau^n)}{\partial \tau^n}\frac{\partial \tau^n}{\partial \ga} \frac{\partial \ga}{\partial e_n} = 1}
  \label{eq:lobtw}
\end{equation}
That is, at this stage, the lobby chooses the level of effort that equates its expected marginal increase in profits with its marginal payment.

@@

\bigskip
\subsection{To Break or Not to Break?}
\label{sec:break}
With trade-war decisions specified, we can proceed by backward induction to analyze the incentives regarding the legislature's decision to uphold or break the trade agreement. I assume that the form of the political weighting function is the same \textbf{regardless of the break decision}, but the model could be extended to allow the two to differ. What is in principle different are the lobbying efforts toward influencing the break decision, denoted $e_b$ at this stage to distinguish them from those that are made to influence the Nash tariff, $e_n$. The realization of uncertainty about the median legislator's identity will similarly be denoted $\ve_b$ so as to distinguish it from $\ve_n$ at the trade-war stage.

@@

The executives' incentive to raise tariffs in order to reduce lobbying effort \textbf{provides insight into the empirical puzzle} in the Protection for Sale literature that levels of protection and associated deadweight losses are too high relative to lobbying expenditure given the high estimates for governments' weighting of social welfare. I have shown that this serves to mediate the relationship between preferences for contributions relative to social welfare and the tariffs that are provided. The intuition is clear: the observed lobbying expenditure levels may in fact be low \textit{because} tariffs have been raised sufficiently high to prevent political pressure and the increased risk of a costly trade disruption it engenders.
